Models

### 2023-01-17 analysis/05_model_metrics.Rmd
Add outlier checks

>>>>

#### Examine Outliers

Let's identify any major outliers in slope random effects for the base models to see if they're impacting the overall trend.

First two functions to visualize and exclude outliers

```{r}
test_normality <- function(model_fit, metric_name){
  intercepts <- ranef(model_fit)$study_id$`(Intercept)`
  slopes <- ranef(model_fit)$study_id$`year_scaled`

  hist(intercepts, main = paste("intercepts for", metric_name))
  hist(slopes, main = paste("slopes for", metric_name))

  bind_rows(tidy(shapiro.test(intercepts)) %>% mutate(type = "intercept"),
            tidy(shapiro.test(slopes)) %>% mutate(type = "slope")) %>%
    mutate(metric = metric_name)
}

check_outliers <- function(start_mod, metric_name, outlier_criterion, lmer_formula, REML = TRUE, bobyqa = FALSE){

  intercepts <- ranef(start_mod)$study_id$`(Intercept)`
  slopes <- ranef(start_mod)$study_id$`year_scaled`

  rows <- which(eval(rlang::parse_expr(outlier_criterion))) # It's row 22
  print(rows)
  if(length(rows) > 1) warning(paste("There is more than one study outside the outlier cutoff for", metric_name))

  study <- as.integer(rownames(ranef(start_mod)$study_id[rows,]))

  if (isTRUE(REML)){
    if (isTRUE(bobyqa)) {
        fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study), 
              control = lmerControl(optimizer = "bobyqa"))
    } else {
      fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study))
    }
  } else{
    if (isTRUE(bobyqa)) {
        fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study), 
              REML = FALSE,
              control = lmerControl(optimizer = "bobyqa"))
    } else {
      fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study),
              REML = FALSE)
    }
  }

  tibble(metric = metric_name,
         old_estimate = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(estimate), 
         old_p = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(p.value), 
         new_estimate = tidy(fit) %>% filter(term == "year_scaled") %>% pull(estimate), 
         new_p = tidy(fit) %>% filter(term == "year_scaled") %>% pull(p.value))

Visualize the random effects for the base models
```{r}
base_mods <- model_table %>% 
  filter(model_id == "base") %>%
  select(metric, model, formula, REML, bobyqa) %>%
  group_by(metric) %>% 
  slice_head() %>%
  ungroup()

mod_normality <- map2_dfr(base_mods$model, base_mods$metric, test_normality)
```

Based on the visualization, we identify some metrics that are suspect for outliers
```{r}
# create dataframe for outlier criterion based on visualize of random effects
criterion <- tibble(metric_name = c("CWM_bodymass_value", "CWM_diet_inv", "CWM_diet_scav", "CWM_diet_vect", 
                                     "CWM_diet_vend", "CWM_diet_vunk", "CWM_forstrat_aerial", "CWM_forstrat_midhigh",
                                     "CWM_forstrat_understory", "Jaccard_base", "S"),
                    outlier_criterion = c("slopes > 0.4", "slopes < -0.5", "slopes > 0.6", "slopes > 0.4", 
                                          "slopes > 0.5", "slopes < -0.5", "slopes > 0.04", "slopes < -0.05", 
                                          "slopes < -0.4", "slopes > 0.02", "slopes < -1")
                    )

# get info for how original models were fit
outlier_mods <- trans_model_df %>% 
  select(start_mod = model, metric_name = metric, lmer_formula = formula, REML, bobyqa) %>%
  right_join(criterion) %>%
  select(start_mod, metric_name, outlier_criterion, lmer_formula, REML, bobyqa)

# refit without outliers and compare
outlier_comp <- pmap_dfr(outlier_mods, check_outliers)
```
